Reasoning about a Capability Machine with Local Capabilities

### 2017-06-07 popl-2018/paper.tex
updated figures to better fit with the new format

% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\decode}{\plainfun{decode}{}}
\newcommand{\encode}{\plainfun{encode}{}}
\newcommand{\encodePerm}{\mathit{encodePerm}}
\newcommand{\encodePermPair}{\plainfun{encodePermPair}{}}
\newcommand{\encodeLoc}{\mathit{encodeLoc}{}}
\newcommand{\decodePermPair}{\plainfun{decodePermPair}}
\newcommand{\decodePerm}[1]{\plainfun{decodePerm}{#1}}
\newcommand{\updatePcPerm}[1]{\plainfun{updatePcPerm}{#1}}

\newcommand{\executeAllowed}[1]{\plainfun{executeAllowed}{#1}}
\newcommand{\nonZero}[1]{\plainfun{nonZero}{#1}}
\newcommand{\readAllowed}[1]{\plainfun{readAllowed}{#1}}
\newcommand{\writeAllowed}[1]{\plainfun{writeAllowed}{#1}}
\newcommand{\withinBounds}[1]{\plainfun{withinBounds}{#1}}
\newcommand{\stdUpdatePc}[1]{\plainfun{updatePc}{#1}}

@@

% Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\Mems}{\Heaps}
%\newcommand{\HeapSegments}{\plaindom{MemSegment}}
\newcommand{\HeapSegments}{\plaindom{MemSeg}}
\newcommand{\MemSegments}{\HeapSegments}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

\begin{figure}
\[
  \begin{array}{rlrcl p{2cm}rlrcl}
    \addr   &\in& \Addrs &\isdef& \nats & & r &\in& \RegName&\bnfdef& \pcreg\mid r_0\mid r_1\mid\ldots\\
    w &\in&\Words &\isdef& \ints + \Caps & & \reg &\in& \Regs  &\isdef& \RegName \rightarrow \Words\\
    \perm   &\in& \Perms &::= & \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid  & & m&\in& \Heaps &\isdef& \Addrs \rightarrow \Words \\
    & & & &\exec\mid \entry\mid \rwx\mid \rwlx & & \Phi &\in& \ExecConfs  &\isdef& \Regs \times \Heaps \\
    \gl&\in&\Globals & ::= &\glob \mid \local & & \ms &\in& \MemSegments &::= & \Addrs \parfun \Words \\
    &&\Confs &::=& \multicolumn{5}{l}{\ExecConfs + \{\failed \} + \{\halted\} \times \Heaps}\\
     && \Caps  &::= & \multicolumn{5}{l}{\left\{
                         ((\perm,\gl),\start,\addrend,\addr)\mid
                         \start,\addr\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}\right\}}
  \end{array}
\]
  \begin{equation*}
  \begin{array}{rcl}
    \rv    &\in& \ints + \RegName \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid \plus{\lv}{\rv}{\rv} \mid 
                 \minus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid \\
           &   & \subseg{\lv}{\rv}{\rv} \mid  
                 \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid
                 \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\end{equation*}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
      \begin{array}[t]{l}
\text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$}\\ \text{\;\;and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }
      \end{array}
\\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                             \text{if $\memreg(\pcreg) = \stdcap$ and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}\\
                             \failed & \text{otherwise}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{3.9cm}|p{8.3cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{w})$&$w = \memreg(r_2)$\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\stdUpdatePc{\updateReg{r_1}{w}}$&$\memreg(r_2) = \stdcap{}$ and $w = \memheap(\addr)$ \newline and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$ \\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$ and $(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      and $(\perm',g') \sqsubseteq (\perm,g)$ and  $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& if $\memreg(r) = ((\entry,\gl),\start,\addrend,\addr)$, then $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$\newline otherwise $\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if $\var{w} = ((\_,\local),\_,\_,\_)$, then $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

@@

\subsection{Logical relation}
\afterpage{
\begin{figure}[htbp]
  \centering
%  \begin{withmathindent}{0cm}
    \begin{align*}
      & \observations \; : \;  \Worlds \nefun \UPred{\Regs \times \HeapSegments} \\
      & \observations (W) \defeq \\
      & \;\left\{ \npair{(\reg,\ms)} \; \middle| \;
        \begin{aligned}
          & \forall \ms_f, \heap', i \leq n \ldotp \\
          & (\reg,\ms \uplus \ms_f) \step[i] (\halted,\heap')  \\
          & \;\; \Rightarrow
          \begin{aligned}[t]
            & \exists W' \futurestr W, \ms_r, \ms' \ldotp\\
            & \;\; \heap' = \ms' \uplus \ms_r \uplus \ms_f \text{ and } \\
            & \;\; \heapSat[\hs']{n-i}{W'}
          \end{aligned}
        \end{aligned}\right\}\\
      & \stdrr \; : \; \Worlds \monwknefun \UPred{\Regs} \\
      & \stdrr(W) \defeq \left\{ \npair{\reg} \; \middle| \;
          \forall r \in \RegName \setminus \{\pcreg\} \ldotp \npair{\reg(r)}
          \in \stdvr(W) \right\} \\
      & \\[-.2cm]
      & \stder \; : \; \Worlds \nefun \UPred{\Words} \\
      & \stder(W) \defeq \left\{ \npair{\pc} \; \middle| \;
        \begin{multlined}
          \forall n' \leq n, \npair[n']{\reg} \in \stdrr(W), \heapSat[\hs]{n'}{W} \ldotp \\
          \npair[n']{(\reg\update{\pcreg}{\pc},\hs)} \in \observations(W)
        \end{multlined} \right\} \\
      & \\
      &\stdvr \; : \; \Worlds \monwknefun \UPred{\Words} \\
      &\stdvr(W)\defeq  \\
      &\;\;\begin{aligned}[t] & \{ \npair{i} \mid i \in \ints \}
        \union \\
        & \{ \npair{\stdcap[(\noperm,\gl)] } \}
        \union \\
        & \left\{
          \begin{multlined}
            \npair{\stdcap[(\readonly,\gl)] } \mid \\
            \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
          \end{multlined} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\readwrite,\gl)] } \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             &\quad\npair{(\start,\addrend)} \in
            \writeCond{}(\iota^\nwl,\gl)(W)
          \end{aligned} \right\} \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\readwritel,\gl)] } \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             &\quad\npair{(\start,\addrend)} \in
            \writeCond{}(\iota^\pwl,\gl)(W)
          \end{aligned} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\exec,\gl)]} \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             &\quad\npair{(\{\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{aligned} \right\}
        \union \\
        & \left\{
          \begin{multlined}
            \npair{\stdcap[(\entry,\gl)]} \mid \\
            \npair{(\start,\addrend,\addr)} \in \entryCond{}(\gl)(W)
          \end{multlined} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\rwx,\gl)]} \mid \\
            &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
            &\quad\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W) \text{ and }\\
            &\quad\npair{(\{\rwx,\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{aligned} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\rwlx,\gl)]} \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             &\quad\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W) \text{ and }\\
             &\quad\npair{(\{\rwlx,\rwx,\exec\},\start,\addrend)} \in
            \execCond{}(\gl)(W)
          \end{aligned} \right\}
      \end{aligned}
    \end{align*}
%  \end{withmathindent}
  \caption{Logical Relation}
  \label{fig:logrel}            
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
  & \readCond{}(\gl)(W) =  \\
  & \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{aligned} \right\}\\
  & \writeCond{}(\iota,\gl)(W) =  \\
  & \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; W(r) \text{ is address-stratified and} \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
  & \execCond{}(\gl)(W) = \\
  & \left\{
    \begin{multlined}
      \npairP{(\var{perms}, \\
        \start,\addrend)}
    \end{multlined}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{perms} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
  & \entryCond{}(\gl)(W) = \\
  & \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{aligned}
 &  \forall n' < n \ldotp \forall W' \future W \ldotp\\
      & \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{aligned} \right\} \\
  & \quad \text{where } \gl = \local \Rightarrow \future = \futurewk \\
  & \quad \text{and } \gl = \glob \Rightarrow \future = \futurestr
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation now defines semantically when values, terms, and
configurations can safely be used or executed by the adversary, i.e. without the
risk that they will be able to use it to break protocols in the world. The
definition of the logical relation is found in \figurename{}s~\ref{fig:logrel}
and~\ref{fig:perm-conds} and we provide some explanations in the following
paragraphs. We omit some definitions and explain them only verbally, but precise
definitions can be found in the technical appendix~\citep{technical_appendix}.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem can be understood as a general expression of the
guarantees offered by the capability machine, an instance of a general property
called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}. To
understand this, consider that the theorem says the capability
$((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter, without any
assumption about what instructions it actually points to (the only assumptions
we have are about the read or write authority that it carries). As such, the
theorem expresses the capability safety of the machine, which guarantees that
\emph{any} instruction is fine and will not be able to go beyond the authority
of the values it has access to.

>>>>

% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\decode}{\plainfun{decode}{}}
\newcommand{\encode}{\plainfun{encode}{}}
\newcommand{\encodePerm}{\mathit{encodePerm}}
\newcommand{\encodePermPair}{\plainfun{encodePermPair}{}}
\newcommand{\encodeLoc}{\mathit{encodeLoc}{}}
\newcommand{\decodePermPair}{\plainfun{decodePermPair}}
\newcommand{\decodePerm}[1]{\plainfun{decodePerm}{#1}}
\newcommand{\updatePcPerm}[1]{\plainfun{updPcPerm}{#1}}

\newcommand{\executeAllowed}[1]{\plainfun{executeAllowed}{#1}}
\newcommand{\nonZero}[1]{\plainfun{nonZero}{#1}}
\newcommand{\readAllowed}[1]{\plainfun{readAllowed}{#1}}
\newcommand{\writeAllowed}[1]{\plainfun{writeAllowed}{#1}}
\newcommand{\withinBounds}[1]{\plainfun{withinBounds}{#1}}
\newcommand{\stdUpdatePc}[1]{\plainfun{updPc}{#1}}

@@

% Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\Mems}{\Heaps}
\newcommand{\HeapSegments}{\plaindom{MemSeg}}
\newcommand{\MemSegments}{\HeapSegments}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

\begin{figure}
\[\arraycolsep=1.4pt
  \begin{array}{rlrcl p{5cm}rlrcl}
    \addr   &\in& \Addrs &\isdef& \nats &\phantom{makesomero}                                & r    &\in & \RegName     &\bnfdef& \pcreg\mid r_0\mid r_1\mid\ldots\\
    w &\in&\Words &\isdef& \ints + \Caps &                                                   & \reg &\in & \Regs        &\isdef& \RegName \rightarrow \Words\\
    \perm   &\in& \Perms &::= & \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid  & & m    &\in & \Heaps       &\isdef& \Addrs \rightarrow \Words \\
    & & & &\exec\mid \entry\mid \rwx\mid \rwlx &                                             & \Phi &\in & \ExecConfs   &\isdef& \Regs \times \Heaps \\
    \gl&\in&\Globals & ::= &\glob \mid \local &                                              & \ms  &\in & \MemSegments &::= & \Addrs \parfun \Words \\
    &&\Confs &::=& \multicolumn{6}{l}{\ExecConfs + \{\failed \} + \{\halted\} \times \Heaps}\\
     && \Caps  &::= & \multicolumn{6}{l}{\left\{
                         ((\perm,\gl),\start,\addrend,\addr)\mid
                         \start,\addr\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}\right\}}
  \end{array}
\]
  \begin{equation*}
  \begin{array}{rcl}
    \rv    &\in& \ints + \RegName \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid \plus{\lv}{\rv}{\rv} \mid 
                 \minus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid \\
          &   &  \restricttwo{\lv}{\rv} \mid
                 \subseg{\lv}{\rv}{\rv} \mid  
                 \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid \\
          &   &  \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\end{equation*}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
      \begin{array}[t]{l}
        \text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$}\\ 
        \text{\;\;and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }
      \end{array}
\\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                           \begin{array}[t]{l}
                             \text{if $\memreg(\pcreg) = \stdcap$}\\
                             \text{\;\;and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}
                           \end{array}
\\
                             \failed & \text{otherwise}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{3.4cm}|p{7.3cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{w})$&$w = \memreg(r_2)$\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\stdUpdatePc{\updateReg{r_1}{w}}$&$\memreg(r_2) = \stdcap{}$ and  $w = \memheap(\addr)$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$ \\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$ and $(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      and $(\perm',g') \sqsubseteq (\perm,g)$ and  $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& if $\memreg(r) = ((\entry,\gl),\start,\addrend,\addr)$, then $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$ otherwise $\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if $\var{w} = ((\_,\local),\_,\_,\_)$, then $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

@@

\subsection{Logical relation}
\afterpage{
\begin{figure}[htbp]
  \centering
%  \begin{withmathindent}{0cm}
    \begin{align*}
      & \observations \; : \;  \Worlds \nefun \UPred{\Regs \times \HeapSegments} \\
      & \observations (W) \defeq 
        \;\left\{ \npair{(\reg,\ms)} \; \middle| \;
        \begin{aligned}
          & \forall \ms_f, \heap', i \leq n \ldotp  (\reg,\ms \uplus \ms_f) \step[i] (\halted,\heap')  \\
          & \;\; \Rightarrow
          \begin{aligned}[t]
            & \exists W' \futurestr W, \ms_r, \ms' \ldotp\\
            & \;\; \heap' = \ms' \uplus \ms_r \uplus \ms_f \text{ and } \heapSat[\hs']{n-i}{W'}
          \end{aligned}
        \end{aligned}\right\}\\
      & \stdrr \; : \; \Worlds \monwknefun \UPred{\Regs} \\
      & \stdrr(W) \defeq \left\{ \npair{\reg} \; \middle| \;
          \forall r \in \RegName \setminus \{\pcreg\} \ldotp \npair{\reg(r)}
          \in \stdvr(W) \right\} \\
      & \\[-.2cm]
      & \stder \; : \; \Worlds \nefun \UPred{\Words} \\
      & \stder(W) \defeq \left\{ \npair{\pc} \; \middle| \;
        \begin{multlined}
          \forall n' \leq n, \npair[n']{\reg} \in \stdrr(W), \heapSat[\hs]{n'}{W} \ldotp \\
          \npair[n']{(\reg\update{\pcreg}{\pc},\hs)} \in \observations(W)
        \end{multlined} \right\} \\
      & \\
      &\stdvr \; : \; \Worlds \monwknefun \UPred{\Words} \\
      &\stdvr(W)\defeq 
        \begin{aligned}[t] & \{ \npair{i} \mid i \in \ints \} \union \{ \npair{\stdcap[(\noperm,\gl)] } \}
        \union \\
        & \left\{
            \npair{\stdcap[(\readonly,\gl)] } \mid \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
          \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\readwrite,\gl)] } \; \middle| \;
            \begin{array}{l}
              \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
              \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
            \end{array}
          \right\} \union \\
        & \left\{
            \npair{\stdcap[(\readwritel,\gl)] } \;\middle| \;
            \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W)
            \end{array}
           \right\}
        \union \\
        & \left\{
          \npair{\stdcap[(\exec,\gl)]} \; \middle| \;
          \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\{\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{array}
           \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\entry,\gl)]} \; \middle| \;
            \npair{(\start,\addrend,\addr)} \in \entryCond{}(\gl)(W)
          \right\}
        \union \\
        & \left\{
          \npair{\stdcap[(\rwx,\gl)]}\; \middle| \; 
          \begin{array}{l}
            \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
            \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W) \text{ and }\\
            \npair{(\{\rwx,\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{array}
          \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\rwlx,\gl)]} \; \middle| \;
            \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W) \text{ and }\\
             \npair{(\{\rwlx,\rwx,\exec\},\start,\addrend)} \in
            \execCond{}(\gl)(W)
            \end{array}
          \right\}
      \end{aligned}
    \end{align*}
%  \end{withmathindent}
  \caption{Logical relation}
  \label{fig:logrel}            
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
  & \readCond{}(\gl)(W) =  \left\{ \npair{(\start,\addrend)} \; \middle| \;
       \exists r \in \var{localityReg}(g,W) \ldotp 
       \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \right\}\\
  & \writeCond{}(\iota,\gl)(W) =  \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp W(r) \text{ is address-stratified and} \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
  & \execCond{}(\gl)(W) = 
    \left\{
    \begin{multlined}
      \npairP{(\var{perms}, \\
        \start,\addrend)}
    \end{multlined}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{perms} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
  & \entryCond{}(\gl)(W) = 
    \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \forall n' < n \ldotp \forall W' \future W \ldotp \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \right\} \\
  & \quad \text{where } \gl = \local \Rightarrow \future = \futurewk \\
  & \quad \text{and } \gl = \glob \Rightarrow \future = \futurestr
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation now defines semantically when values, terms, and
configurations can safely be used or executed by the adversary, i.e. without the
risk that they will be able to use it to break protocols in the world. The
definition of the logical relation is found in \figurename{}s~\ref{fig:logrel}
and~\ref{fig:perm-conds} and we provide some explanations in the following
paragraphs. We omit some definitions and explain them only verbally, but precise
definitions can be found in the technical appendix~\citep{technical_appendix}.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalizes the guarantees offered by
the capability machine. Essentially, it says a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{ and } \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{ and }\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and }\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem can be understood as a general expression of the
guarantees offered by the capability machine, an instance of a general property
called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}. To
understand this, consider that the theorem says the capability
$((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter, without any
assumption about what instructions it actually points to (the only assumptions
we have are about the read or write authority that it carries). As such, the
theorem expresses the capability safety of the machine, which guarantees that
\emph{any} instruction is fine and will not be able to go beyond the authority
of the values it has access to.
